---
title: "Examples of Samplers"
author: "Shunri Zheng"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(HDInterval)
library(invgamma)
library(combinat)
library(MASS)
library(MCMCpack)
library(emdbook)
library(mvtnorm)
```




```{r}
set.seed(20230225)


# Define the mixture parameters
p1 <- 1/5
p2 <- 4/5
mu1 <- c(-5, -5)
mu2 <- c(5, 5)
sigma <- diag(2)

# Define the PDF function
pdf_mixture <- function(x, p1, p2, mu1, mu2, sigma) {
  dens1 <- p1 * dmvnorm(x, mean = mu1, sigma = sigma)
  dens2 <- p2 * dmvnorm(x, mean = mu2, sigma = sigma)
  return(dens1 + dens2)
}

gradient_mixture <- function(x, p1, p2, mu1, mu2, sigma){
  dens1 <- p1 * dmvnorm(x, mean = mu1, sigma = sigma)
  dens2 <- p2 * dmvnorm(x, mean = mu2, sigma = sigma)
  numer <- dens1 + dens2
  grad_x1 <- (p1 * dmvnorm(x, mean = mu1, sigma = sigma) / numer) *(-0.5) * solve(sigma) %*% (x - mu1)
  grad_x2 <- (p2 * dmvnorm(x, mean = mu2, sigma = sigma) / numer) *(-0.5) * solve(sigma) %*% (x - mu2)
return(grad_x1+grad_x2)
}

logpdf<-function(x){
  log(pdf_mixture(x, p1, p2, mu1, mu2, sigma))
}


pdf_mixture(c(5,5), p1, p2, mu1, mu2, sigma)
gradient_mixture(c(5,4), p1, p2, mu1, mu2, sigma)

```



### Gibbs Sampler Implementation

Gibbs sampler for (U,V) is implemented as follows:

```{r}
set.seed(20230225)

x_given_y <- function(y, p1, p2, mu1, mu2, sigma) {
    if (runif(1) < p1 / (p1 + p2)) {
        return(rnorm(1, mean = mu1[1]+ sigma[1,2] / sigma[1,1] * (y - mu1[2]), sd = sqrt(sigma[1,1] - sigma[1,2]^2 / sigma[2,2])))
    } else {
        return(rnorm(1, mean = mu2[1]+ sigma[1,2] / sigma[1,1] * (y - mu2[2]), sd = sqrt(sigma[1,1] - sigma[1,2]^2 / sigma[2,2])))
    }
}

y_given_x  <- function(x, p1, p2, mu1, mu2, sigma){
    if (runif(1) < p1 / (p1 + p2)) {
        return(rnorm(1, mean = mu1[2]+ sigma[2,1] / sigma[2,2] * (x - mu1[1]), sd = sqrt(sigma[2,2] - sigma[2,1]^2 / sigma[1,1])))
    } else {
        return(rnorm(1, mean = mu2[2]+ sigma[2,1] / sigma[2,2] * (x - mu2[1]), sd = sqrt(sigma[2,2] - sigma[2,1]^2 / sigma[1,1])))
    }
}

sampleGibbs2 <- function(start.x, start.y, c, n.sims, burnin){

    # initialize the chain
    chain <- matrix(NA, nrow=n.sims, ncol=2)
    chain[1,] <- c(start.x, start.y)

    # loop through the chain
    for(i in 2:n.sims){
        chain[i,1] <- x_given_y(chain[i-1,2], p1, p2, mu1, mu2, sigma)
        chain[i,2] <- y_given_x(chain[i,1], p1, p2, mu1, mu2, sigma)
    }

    # return the chain
    return(chain[(burnin+1):n.sims,])
}

Samples_From_Gibbs1 = sampleGibbs2(0,0,0.25,5000,1000)
```

### Traceplots and Scatterplots

```{r}
ts.plot(Samples_From_Gibbs1[,1], type="l", col="black", lwd=1, xlab="Iteration",
 ylab="x", main="Trace plot of x")
ts.plot(Samples_From_Gibbs1[,2], type="l", col="black", lwd=1, xlab="Iteration",
ylab="y", main="Trace plot of y")
ggplot(data.frame(Samples_From_Gibbs1), aes(x=Samples_From_Gibbs1[,1], y=Samples_From_Gibbs1[,2])) + geom_point(alpha = 0.5,size = 0.5,color = "blue") + ggtitle("Scatterplot of (x,y)")
```
```{r}
mean(Samples_From_Gibbs1[,1])
mean(Samples_From_Gibbs1[,2])
```


### Metropolis-Hastings Sampler Implementation

Metropolis-Hastings sampler for (X,Y) is implemented as follows:

```{r}

density <- function(x,y,c){
    if(abs(x-y) <= c & x>=0 & x<=1 & y>=0 & y<=1){
        return(1)
    }else{
        return(0)
    }
}

```

```{r}
# density function for mixed normal distribution
density <- function(x,y,c){
    if(abs(x-y) <= c & x>=0 & x<=1 & y>=0 & y<=1){
        return(1)
    }else{
        return(0)
    }
}

```


```{r}
set.seed(20230225)
sampleL <- function(start.a, start.b,c, n.sims,burnin=0){}

```







```{r}


set.seed(20230225)
sampleMH <- function(start.a, start.b, n.sims,burnin=0){
# initialize the chain
chain <- matrix(NA, nrow=n.sims, ncol=2)
chain[1,] <- c(start.a, start.b)
accept = 0
# loop through the chain    
for(i in 2:n.sims){
    noise = rnorm(2,0,5)
    new.a <- chain[i-1,] + noise
    if(pdf_mixture(new.a,p1,p2,mu1,mu2,sigma) <= 0){
        chain[i,] <- chain[i-1,]
    } else{
        ratio <- pdf_mixture(new.a,p1,p2,mu1,mu2,sigma)/pdf_mixture(chain[i-1,],p1,p2,mu1,mu2,sigma)
    if(runif(1) < ratio){
        chain[i,] <- new.a
        accept = accept+1
    }else{
        chain[i,] <- chain[i-1,]
    }    
}
}
# return the chain
print(accept/(n.sims-1))
return(chain[(burnin+1):n.sims,])
}

Samples_From_MH = sampleMH(0,0,11000,1000)

```


### Traceplots and Scatterplots

```{r}
ts.plot(Samples_From_MH[,1], type="l", col="black", lwd=1, xlab="Iteration",
 ylab="u", main="Trace plot of u for c=0.25")
ts.plot(Samples_From_MH[,2], type="l", col="black", lwd=1, xlab="Iteration",
ylab="v", main="Trace plot of v for c=0.25")
ggplot(data.frame(Samples_From_MH), aes(x=Samples_From_MH[,1], y=Samples_From_MH[,2])) + geom_point(alpha = 0.5,size = 0.5,color = "blue") + ggtitle("Scatterplot of (x,y) using Metropolis-Hastings")
```

```{r}
print(accept)
```





### Langevin Sampler Implementation

Langevin sampler for mixture gaussian


```{r}
set.seed(20230225)
sampleL <- function(start.a, start.b, n.sims,burnin=0,stepsize=0.1){
# initialize the chain
chain <- matrix(NA, nrow=n.sims, ncol=2)
chain[1,] <- c(start.a, start.b)


# loop through the chain
for(i in 2:n.sims){
    noise <- rnorm(2, mean = 0, sd = sqrt(2 * stepsize))
    chain[i,] <- chain[i-1,] + stepsize * gradient_mixture(chain[i-1,],p1,p2,mu1,mu2,sigma) + noise  
}
# return the chain
return(chain[(burnin+1):n.sims,])
}

Samples_From_Langevin = sampleL(-1,0,11000,1000,0.1)


ts.plot(Samples_From_Langevin[,1], type="l", col="black", lwd=1, xlab="Iteration",
 ylab="u", main="Trace plot of u for c=0.25")
ts.plot(Samples_From_Langevin[,2], type="l", col="black", lwd=1, xlab="Iteration",
ylab="v", main="Trace plot of v for c=0.25")
ggplot(data.frame(Samples_From_Langevin), aes(x=Samples_From_Langevin[,1], y=Samples_From_Langevin[,2])) + geom_point(alpha = 0.5,size = 0.5,color = "blue") + ggtitle("Scatterplot of (x,y) using Langevin Sampler")
```


### Traceplots and Scatterplots

```{r}

```



### Metropolis-adjusted Langevin algorithm Implementation

```{r}
set.seed(20230225)

q <- function(x,y,p1,p2,mu1,mu2,sigma,stepsize){
    return (exp(-1/(4*stepsize)*sum((x-y-stepsize*gradient_mixture(y,p1,p2,mu1,mu2,sigma))**2)))
}





sampleMALA <- function(start.a, start.b, n.sims,burnin=0,stepsize=0.1){
# initialize the chain
chain <- matrix(NA, nrow=n.sims, ncol=2)
chain[1,] <- c(start.a, start.b)


# loop through the chain
for(i in 2:n.sims){
    noise <- rnorm(2, mean = 0, sd = sqrt(2 * stepsize))
    x_new <-  as.vector(chain[i-1,] + stepsize * gradient_mixture(chain[i-1,],p1,p2,mu1,mu2,sigma) + noise)
    ratio <- pdf_mixture(x_new,p1,p2,mu1,mu2,sigma)*q(chain[i-1,],x_new,p1,p2,mu1,mu2,sigma,stepsize)/(pdf_mixture(chain[i-1,],p1,p2,mu1,mu2,sigma)*q(x_new,chain[i-1,],p1,p2,mu1,mu2,sigma,stepsize))
    if(runif(1) < ratio){
        chain[i,] <- x_new
}else{
    chain[i,] <- chain[i-1,]
}
}
# return the chain
return(chain[(burnin+1):n.sims,])
}
Samples_From_MALA = sampleMALA(0,0,11000,1000,0.1)

```

### Traceplots and Scatterplots

```{r}

ts.plot(Samples_From_MALA[,1], type="l", col="black", lwd=1, xlab="Iteration",
 ylab="u", main="Trace plot of u for c=0.25")
ts.plot(Samples_From_MALA[,2], type="l", col="black", lwd=1, xlab="Iteration",
ylab="v", main="Trace plot of v for c=0.25")
ggplot(data.frame(Samples_From_MALA), aes(x=Samples_From_MALA[,1], y=Samples_From_MALA[,2])) + geom_point(alpha = 0.5,size = 0.5,color = "blue") + ggtitle("Scatterplot of (x,y) using MALA")
```
```

### Hamiltonian Monte Carlo Implementation

```{r}
```
